---
title: "Final_project"
author: "Mavis Fosuaa Boateng"
date: "2024-07-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

This is the my final project work. It entails  processes of data importation,  process of cleaning and analyzing tuberculosis data across various countries and years. This project will explore trends and patterns.

## Objectives

-   Pivot the tuberculosis datasets to a long format.
-   Handle non-numeric values in case columns that include characters.
-   Standardize country names using ISO codes to facilitate data merging.
-   Join the cleaned tuberculosis data with population data from the {tidyr} package.
-   Visualize the data to analyze trends.



```{r}
# Load packages 
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse, gapminder,
               here,visdat, 
               inspectdf,
               countrycode,   bslib, bsicons, shiny, 
               rnaturalearth, plotly, htmltools, 
               reactable,sf,ggplot2,sf, 
               rgeoboundaries
               )

```

# Data Importation

Raw dataset is about:

1.The total number of cases of tuberculosis per 100,000 people.It also captures the proportion of estimated new and relapse TB cases that were notified  and treated by health facilities in the country during the given year.


```{r}
TB_old_cases <- read.csv(here("data/sh_tbs_dtec_zs.csv"))

TB_new_cases<-read.csv(here("data/sh_tbs_incd.csv"))

TB_deaths<-read.csv(here("data/all_forms_of_tb_deaths_per_100000_estimated.csv"))

TB_treatment<-read.csv(here("data/sh_tbs_cure_zs.csv"))
```



```{r}
# Show the first 6 rows
head(TB_old_cases)
str(TB_old_cases)

head(TB_new_cases)
str(TB_new_cases)

head(TB_deaths)
str(TB_deaths)

```




```{r}
inspect_cat(TB_old_cases) %>%
 show_plot()

inspect_cat(TB_new_cases) %>%
 show_plot()
```

# Data Cleaning

## Pivot to Long Format

Converting the datasets from wide to long format will make merging easier.

#Data cleaning for old TB Cases
```{r long-format-total}
# Filter columns to only include the years 2010 to 2020
TB_cases_filtered  <- TB_old_cases %>%
  select(country, `X2000`: `X2020`)

# Converting from wide to long format
TB_cases_long <-  TB_cases_filtered%>%
  pivot_longer(
    cols = -country,       # Exclude the 'country' column
    names_to = "year",     # Name of the new 'year' column
    values_to = "cases" # Name of the new 'total_cases' column
  )

# View the long format data
print(TB_cases_long)



# Remove 'X' from the 'year' column
TB_cases_long$year <- gsub('X', '', TB_cases_long$year)

# Convert 'year' column to numeric
TB_cases_long$year <- as.numeric(TB_cases_long$year)

# View the updated dataframe
print(TB_cases_long)

# Remove the decimal point and convert back to numeric
TB_cases_long$cases <- as.numeric(as.integer(TB_cases_long$cases))

# Print the updated dataframe
print(TB_cases_long)

```




## Converting the TB_new_cases dataset from wide to long format and other data cleaning


```{r long-format-total}
# Filter columns to only include the years 2010 to 2020
TB_new_cases_filtered  <- TB_new_cases %>%
  select(country, `X2000`: `X2020`)

# Converting from wide to long format
TB_new_long <-  TB_new_cases_filtered%>%
  pivot_longer(
    cols = -country,       # Exclude the 'country' column
    names_to = "year",     # Name of the new 'year' column
    values_to = "new_cases" # Name of the new 'total_cases' column
  )

# View the long format data
print(TB_new_long)


# Remove 'X' from the 'year' column
TB_new_long$year <- gsub('X', '', TB_new_long$year)

# Convert 'year' column to numeric
TB_new_long$year <- as.numeric(TB_new_long$year)
# View the updated dataframe
print(TB_new_long)



# Convert data types from chr to num

TB_new_long <- TB_new_long %>%
  mutate(
    # Convert year to numeric
    year = as.numeric(year),
    # Remove 'k' and 'M' and convert to numeric
    new_cases= case_when(
      str_detect(new_cases, 'k$') ~ as.numeric(str_remove(new_cases, 'k')) * 1000,
      str_detect(new_cases, 'M$') ~ as.numeric(str_remove(new_cases, 'M')) * 1000000,
      TRUE ~ as.numeric(new_cases)
    ))

TB_new_long$new_cases <- as.numeric(as.integer(TB_new_long$new_cases))

```

## Data cleaning for number of TB deaths.Converting the dataset from wide to long format.


```{r long-format-total}
# Filter columns to only include the years 2010 to 2020
TB_deaths_filtered  <- TB_deaths %>%
  select(country, `X2000`: `X2020`)

# Converting from wide to long format
TB_mortality <-  TB_deaths_filtered%>%
  pivot_longer(
    cols = -country,       # Exclude the 'country' column
    names_to = "year",     # Name of the new 'year' column
    values_to = "deaths" # Name of the new 'total_cases' column
  )

# View the long format data
print(TB_mortality)


# Remove 'X' from the 'year' column
TB_mortality$year <- gsub('X', '', TB_mortality$year)

# Convert 'year' column to numeric
TB_mortality$year <- as.numeric(TB_mortality$year)

# View the updated dataframe
print(TB_mortality)

# Remove the decimal point and convert back to numeric
TB_mortality$deaths <- as.numeric(as.integer(TB_mortality$deaths))

# Print the updated dataframe
print(TB_mortality)

```

#Data cleaning for TB treatment data
```{r long-format-total}
# Filter columns to only include the years 2010 to 2020
TB_treatment_filtered  <- TB_treatment %>%
  select(country, `X2000`: `X2020`)

# Converting from wide to long format
TB_treatment <-  TB_treatment_filtered%>%
  pivot_longer(
    cols = -country,       # Exclude the 'country' column
    names_to = "year",     # Name of the new 'year' column
    values_to = "treatment" # Name of the new 'total_cases' column
  )

# Remove 'X' from the 'year' column
TB_treatment$year <- gsub('X', '', TB_treatment$year)

# Convert 'year' column to numeric
TB_treatment$year <- as.numeric(TB_treatment$year)

# Print the updated dataframe
print(TB_treatment)

```
## Joining Datasets

Join the datasets using the common keys, `country` and `year`.

```{r join-datasets}

# Perform the left join
TB_cases_combined <- left_join(TB_cases_long,TB_new_long,
                              by = c("country", "year"))

TB_dat <- left_join(TB_cases_combined,TB_mortality,
                           by = c("country", "year"))

TB_data <- left_join(TB_dat,TB_treatment,
                           by = c("country", "year"))
# View the combined dataset
print(TB_data)

```



```{r}
TB_dataset <- TB_data %>%
  mutate(total_cases = cases+ new_cases)
print(TB_dataset)

```


# Line graphs showing the confirmed cases of TB cases among three selected countries 

```{r}
# Load necessary libraries
library(ggplot2)
 library(dplyr)

# Correct number of repetitions for the 'country' column
countries <- rep(c("Zimbabwe", "South Africa", "India"), each = 11)

# Correct number of years
years <- rep(2010:2020, times = 3)


# List of countries to plot
country_list <- c("Zimbabwe", "South Africa", "India")

# Loop through each country and create the plot
for (i in country_list) {
  # Subset data for the current country
  country_data <- subset(TB_dataset, country == i)
  
  # Create the plot
  plot_1 <- ggplot(country_data, aes(x = year, y = total_cases)) +
    geom_line(color = "steelblue", size = 1) +          # Line color and size
    geom_point(color = "darkred", size = 3) +           # Points color and size
    labs(
      title = paste("TB Cases in", i),
      x = "Year",
      y = "total_cases"
    ) +
    theme_minimal(base_size = 15) +                     # Base font size for theme
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 20),  # Center the title
      axis.title.x = element_text(size = 14, face = "bold"),             # X-axis title
      axis.title.y = element_text(size = 14, face = "bold"),             # Y-axis title
      axis.text.x = element_text(angle = 45, hjust = 1),                 # X-axis text
      panel.grid.major = element_line(color = "gray", size = 0.5),       # Major grid lines
      panel.grid.minor = element_line(color = "lightgray", size = 0.25)  # Minor grid lines
    )+
  scale_color_brewer(palette = "Set1")    # Color palette for the lines
  
  # Print the plot
  print(plot_1)
}

```




## Standardizing Country Names

Converting Country Names to ISO Codes: You can use the `countrycode()` function to map your dataset's country names to their corresponding ISO codes.

For example here I add ISO codes to each dataset for a consistent joining key.

```{r add-iso}
TB_dataset <- TB_dataset %>%
  mutate(country_iso = countrycode(country, 
                                   "country.name", 
                                   "iso3c"))
```

```{r}
# Save the dataframe to a CSV file
write.csv(TB_dataset, file = "TB.dataset.csv", row.names = FALSE)

```

# Adding population data


```{r}
pop <- tidyr::population
 pop

 
```





Now, do the same for the `pop` data frame:

```{r}
# Add 3- letter ISO code to pop
pop_iso <- pop %>%
mutate(country_iso = countrycode(country, 
                                   "country.name", 
                                   "iso3c"))
```


## Joining with Population Data


```{r population-join}
# Ensure the 'year' column in both datasets is of the same type
pop_iso <- pop_iso %>%
  mutate(year = as.character(year))  # Convert 'year' to character


TB_dataset <- TB_dataset %>%
  mutate(year = as.character(year))  # Convert 'year' to character

# Join on ISO code and year
data_TB <-
  left_join(pop_iso,TB_dataset,
by =c("country_iso","year") )

data_TB <- data_TB%>% rename(country = country.x)  
  
```





# Mapping country borders with `{rnaturalearth}`

• Let's draw a world map with country borders.

• `{rnaturalearth}` can map all the countries in the world, among others.

• Use `ne_countries()` with the `returnclass = "sf"` argument.



```{r}
# Load the map data
countries <- ne_countries(scale="medium", returnclass = "sf")

# Filter the data for the latest year available
latest_year <- data_TB %>%
  filter(!is.na(year)) %>%
  summarise(latest_year = max(as.numeric(year), na.rm = TRUE)) %>%
  pull(latest_year)

latest_data <- data_TB %>%
  filter(year == latest_year)
world_data <- countries %>%
  left_join(latest_data, by = c("iso_a3" = "country_iso"))

# Create the choropleth map
ggplot(data = world_data) +
  geom_sf(aes(fill =total_cases ), color = "darkblue", size = 0.1) +
  scale_fill_viridis_c(option = "plasma", trans = "log10", na.value = "grey", 
                       name = "TB_cases") +
  theme_minimal() +
  labs(
    title = paste("Global Tuberculosis prevalence in", latest_year),
    subtitle = "Choropleth map showing Tuberculosis cases per 100k population",
    caption = "Source: Tuberculosis data and rnaturalearth dataset"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5)
  )

```



